Step by Step: Using Step Selection Functions to Simulate Dispersal and Assess Landscape Connectivity

### 2021-09-24 Chapter_1/04_Manuscript/Manuscript.tex
Worked on Dom's comments

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
% \usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}

@@

%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
% Importance of dispersal & connectivity and mow its measured
Dispersal of individuals is a crucial prerequisite for long-term species
persistence but depends on a sufficient degree of landscape connectivity. To
date, connectivity has primarily been quantified using least-cost analysis and
circuit theory, albeit both methods make assumptions that are rarely met in
reality. Least-cost analysis assumes that animals move towards a preconceived
destination and choose a cost-minimizing route accordingly, whereas circuit
theory presumes a complete random walk without directional persistence. While
these shortcomings can be overcome by spatio-temporally explicitly simulating
dispersal trajectories from individual-based movement models, a unified
approach for such simulations is lacking.

% What we do
Here, we present a simple three-step workflow to simulate dispersal and asses
landscape connectivity starting from empirical GPS movement data. We exemplify
the workflow using dispersal data collected on the endangered African wild dog
(\textit{Lycaon pictus}) in the Kavango-Zambezi Transfrontier Conservation Area,
the world's largest transboundary conservation area. In step one, we use
integrated step selection functions to parametrize a mechanistic movement model
describing habitat and movement preferences of dispersers. In step two, we
employ the parametrized model to simulate individual dispersal trajectories
across the landscape. In step three, we interpret simulated trajectories by
converting them into three complementary connectivity maps: a heatmap revealing
frequently traversed areas, a betweenness-map delineating dispersal corridors,
and a map of inter-patch connectivity depicting the presence and intensity of
functional links between habitat patches.

% Some results
We find that simulating dispersal from indvidual-based movement models fitted
via step selection functions are readily interpretable and offer great
flexibility to realistically render dispersal movements. Moreover, by not
enforcing connections towards known endpoints but instead letting dispersers
react to the environment as they go, such simulations overcome shortcomings
inherent least-cost path analysis, while at the same time allowing for
autocorrelation in simulated steps, which can not be incorporated using circuit
theory. Inspite of these advantages, simulating dispersal entails numerous
non-trivial modeling decisions, such as the number of simulated individuals and
the duration of simulated dispersal events.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, rescue non-viable populations \citep{Brown.1977}, and colonize
or recolonize unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. However,
the ability to disperse is contingent on a sufficient degree of landscape
connectivity \citep{Fahrig.2003, Clobert.2012}, which is why the identification
and protection of dispersal corridors that promote connectivity has become a
task of fundamental importance \citep{Nathan.2008, Doerr.2011, Rudnick.2012}.
The identification of dispersal corridors not only necessitates a comprehensive
understanding of the factors that limit dispersal, but also an appropriate model
to estimate connectivity \citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To
date, the two most prominent connectivity models in the literature are
least-cost path analysis (LCPA; \citealp{Adriaensen.2003}) and circuit theory
(CT; \citealp{McRae.2006, McRae.2008}), both graph-based approaches that
quantify conductance of the landscape based on habitat permeability
\citep{Zeller.2012}.

% Issues with Both Methods
\subsection{Issues with Traditional Connectivity Models}
Despite their intuitive nature and ease of use, both methods make rigorous
assumptions that are hardly ever met in reality \citep{Diniz.2019}. With LCPA,
for instance, a least-costly route always exists, even if associated movement
costs are unreasonably high and will never be incurred by dispersing individuals
\citep{Sawyer.2011}. The method also presumes that animals move towards a
preconceived endpoint and choose a cost-minimizing route accordingly
\citep{Abrahms.2017}. While these assumptions may be fulfilled by migrating
animals, they are unlikely to hold for dispersers, as dispersers typically move
across unfamiliar territory and are therefore less aware of associated movement
costs \citep{Koen.2014, Cozzi.2020}. CT relaxes these assumptions, but posits
that animals move according to a random walk, entailing that autocorrelation
between subsequent movements cannot be rendered \citep{Diniz.2019}. In reality,
however, autocorrelation in animal movements is regularly observed
\citep{Bovet.1991, Schultz.2001}, especially during dispersal \citep{Cozzi.2020,
Hofmann.2021}. Some of these deficiencies can be alleviated using less stringent
graph-based methods, such as least-cost \textit{corridors} \citep{Pinto.2009},
\textit{thresholded} least-cost paths \citep{Landguth.2012}, or
\textit{randomized} least-cost paths \citep{Panzacchi.2016}, yet a certain
degree of arbitrariness remains. Besides this, graph-based methods are incapable
of rendering the temporal dimension of dispersal, meaning that statements about
the approximate duration required to move between habitats are impossible
\citep{Martensen.2017, Diniz.2019}.

% What about IBMMs?
\subsection{What about IBMMs?}
To overcome the unrealistic assumptions of LCPA and CT, dispersal simulations
from individual-based movement models (IBMMs) have been proposed and applied
\citep{Diniz.2019}. In these models, dispersal trajectories are simulated based
on movement rules that govern how individuals move across and interact with the
prevailing landscape (e.g. \citealp{Gustafson.1996, Gardner.2004, Graf.2007,
KramerSchadt.2004, Revilla.2004, Revilla.2008, Kanagaraj.2013, Clark.2015,
Allen.2016, Hauenstein.2019, Zeller.2020, Vasudev.2021}). Once the trajectories
are simulated, they are further processed and converted into measures of
connectivity \citep{Kanagaraj.2013, Zeller.2020}. Because IBMMs do not enforce
connections towards preconceived endpoints, simulated individuals are allowed to
dynamically adjust their behavior as they disperse. This enables to render
potential interactions between movement behavior and habitat conditions,
shifting the focus from a structural to a more functional view on connectivity
\citep{Tischendorf.2000, Kanagaraj.2013, Hauenstein.2019}. Furthermore, IBMMs
render movement sequentially, meaning that the temporal dimension of dispersal
becomes explicit and that autocorrelation between movements can be incorporated
\citep{Diniz.2019}. In spite of these advantages, a unified framework to
simulate dispersal and assess connectivity using IBMMs is lacking.

% Proposed Solution: Three-Step Workflow
\subsection{Proposed Solution: Three-Step Workflow}
In this paper, we draw on recent innovations in the field of movement ecology
and network theory to propose a simple three-step workflow for the simulation of
dispersal and assessment of connectivity (\Cref{GraphicalAbstract}). In the
first step, GPS data of the focal species is combined with relevant habitat
covariates and analyzed in an integrated step selection framework to fit a
movement model. In the second step, the parametrized model is employed as an
IBMM to simulate dispersal trajectories across the study area. Finally, in the
third step, simulated trajectories are converted into a set of three
complementary connectivity maps, each focusing on a different aspect of
connectivity. The set of maps includes a heatmap, revealing frequently traversed
areas, a betweenness-map, delineating dispersal corridors and bottlenecks, and a
map of inter-patch connectivity, depicting the presence or absence, intensity,
and dispersal duration of functional links between habitat patches.

% Some Background on SSFs
\subsection{Some Background on SSFs}
To fit a movement model, we propose to employ the framework of step selection
functions (SSFs; \citealp{Fortin.2005, Thurfjell.2014, Avgar.2016}). Originally,
SSFs were developed to learn about habitat preferences of the focal species
using GPS data \citep{Fortin.2005}. The method works by comparing covariates
along realized steps (the straight-line connecting two consecutive GPS
relocations; \citealp{Turchin.1998}) to the same covariates along randomly
generated alternative steps \citep{Fortin.2005, Thurfjell.2014} using
conditional logistic regression \citep{Fortin.2005, Muff.2020}. Recently,
\cite{Avgar.2016} have introduced \textit{integrated} SSFs (ISSFs), which is a
generalization of regular SSFs and requires that movement metrics are included
in the corresponding regression model. This not only reduces biases in estimated
preferences, but also permits simultaneous inference on a species' habitat
kernel (i.e. habitat preferences), its movement kernel (i.e. movement
preferences/capabilities), and potential interactions between the two kernels
\citep{Avgar.2016, Fieberg.2020}. Importantly, a model that was fitted using
ISSFs resembles a fully mechanistic IBMM, based on which movement can be
simulated \citep{Avgar.2016, Signer.2017}. Previously, ISSFs have been applied
to simulate utilization distributions of resident animals \citep{Avgar.2016,
Signer.2017}, yet the method remains underutilized for the simulation of
dispersal and detection of movement corridors.

% Some Background on Connectivity "Metrics"
\subsection{Some Background on  Connectivity ``Metrics'' }
Irrespective of the chosen modeling framework, simulated dispersal trajectories
need to be processed into measures of connectivity. To date, the most common
practice to translate simulations into connectivity measures is to compute
inter-patch connectivity between distinct habitats \citep{Diniz.2019}.
Inter-patch connectivity is a simple ratio between the number of dispersers
successfully moving between habitat patches and the total number of simulated
individuals from each patch (e.g. \citealp{Kanagaraj.2013}). As an alternative,
some researchers have tallied overlapping trajectories into what might be called
``heatmaps'', i.e. maps that depict the frequency at which each spatial unit in
the study area was traversed by simulated dispersers (e.g. \citealp{Peer.2008,
Hauenstein.2019, Zeller.2020}). More recently, \cite{BastilleRousseau.2018}
proposed a method to convert trajectories into a network, so that metrics from
network theory can be computed. With regards to connectivity, the betweenness
metric seems promising, as it indicates the number of shortest paths going
through each node in the network and therefore highlights areas of exceptional
importance for connecting different regions in the study area. Because each of
the three metrics metrics focuses on a very different aspect of landscape
connectivity, we propose to view them as complementary and to use them in
concert when assessing connectivity.

@@

% Case Study
\subsection{Case Study}
To exemplify the application of the proposed workflow in
(\Cref{GraphicalAbstract}), we investigate landscape connectivity for the
endangered African wild dog (\textit{Lycon pictus}) as a case study. While once
present across entire sub-Saharan Africa, this species has disappeared from a
vast majority of its historic range, mainly due to persecution by humans,
habitat fragmentation and destruction, and deadly diseases
\citep{Woodroffe.2012}. Some of the biggest remaining populations reside near
the Moremi Game Reserve in northern Botswana \citep{McNutt.1996, Cozzi.2013,
Behr.2020}, which is part of the Kavango-Zambezi Transfrontier Conservation Area
(KAZA-TFCA), the world's largest transboundary protected area. We previously
assessed landscape connectivity for dispersing African wild dogs within this
ecosystem using least-cost methods \citep{Hofmann.2021}. More specifically, we
used ISSFs to fit a habitat selection model and predicted a permeability surface
based on which we computed least-cost paths and corridors between protected
areas. We now expand on this knowledge and use ISSFs to develop a more detailed,
mechanistic movement model of dispersing wild dogs (\Cref{GraphicalAbstract}).
We then employ the model to simulate 80'000 dispersers moving 2'000 steps across
the landscape of the KAZA-TFCA. Finally, we condense the simulated trajectories
into a heatmap, betweenness-map, and inter-patch connectivity map. We thereby
demonstrate that simulations from ISSFs provide a viable alternative to LCPA or
CT and deliver a more comprehensive understanding of the emergence of
connectivity. Nevertheless, we recognize that simulations from ISSFs entail
several non-trivial modeling-decisions, such as the number of simulated
dispersers and the duration of simulated dispersal events.

\section{Methods}
\subsection{Study Area}
The study area of our wild dog case study was defined by a bounding box centered
at -17\degree 13'9''S, 23\degree 56'4''E (\Cref{StudyArea}a) and stretched over
1.3 Mio. km\textsuperscript{2}, encompassing the entire KAZA-TFCA
(\Cref{StudyArea}b). The KAZA-TFCA comprises parts of Angola, Botswana, Namibia,
Zimbabwe, and Zambia and hosts diverse landscapes, ranging from savannah to
grassland and from dry to moist woodland habitats. In its center lies the
Okavango Delta, a dominant hydrogeographical feature and the world's largest
flood-pulsing inland delta. Although large portions of the KAZA-TFCA are
designated national parks and other protected areas, considerable human
influence remains from roads, agricultural sites and settlements.

\begin{figure}[h]
  \begin{center}
    \begin{tikzpicture}
        \node[anchor=south west,inner sep=0] (image) at (0,0,0) {
        \includegraphics[width=\textwidth]{99_StudyArea.png}
        };
        \begin{scope}[x={(image.south east)},y={(image.north west)}]
            % % next four lines will help you to locate the point needed by forming a grid.
            % % comment these four lines in the final picture.
            % \draw[help lines,xstep=.1,ystep=.1] (0,0) grid (1,1);
            % \draw[help lines,xstep=.05,ystep=.05] (0,0) grid (1,1);
            % \foreach \x in {0,1,...,9} { \node [anchor=north] at (\x/10,0) {0.\x}; }
            % \foreach \y in {0,1,...,9} { \node [anchor=east] at (0,\y/10) {0.\y};}
            % % upto here
            \draw[densely dotted, black] (0.169, 0.222) -- (0.364, 0.955);
            \draw[densely dotted, black] (0.169, 0.157) -- (0.364, 0.074);
        \end{scope}
    \end{tikzpicture}
    \caption{Illustration of the study area located in southern Africa. (a) The
    study area was confined by a bounding box spanning the entire KAZA-TFCA and
    comprised parts of Angola, Namibia, Botswana, Zimbabwe, and Zambia. (b)
    The KAZA-TFCA represents the world's largest terrestrial conservation area
    and covers a total of 520'000 km\textsuperscript{2}. Its main purpose is to
    re-establish connectivity between already-existing national parks (dark
    green) and other protected areas (light green).}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{GPS Data}
GPS data of 16 dispersing wild dogs was obtained as described in
\cite{Cozzi.2020} and \cite{Hofmann.2021}. We collected data on 7 female and 9
male dispersers between 2011 and 2019 from a free-ranging population inhabiting
the Moremi National Park in northern Botswana. Because behavior during dispersal
is more pertinent to landscape connectivity than behavior during residence
\citep{Elliot.2014, Abrahms.2017}, we discarded all data that was collected
during residency. For some individuals, exact dispersal dates were known from
field observations, whereas in other cases, we determined dispersal phases using
net-squared displacement. Net-squared displacement is a metric that measures the
squared Euclidean distance of a GPS relocation to a reference point
\citep{Borger.2012}, which in our case was set to the center of each
individual's natal home range. Thus, dispersal was deemed to have started when
an individual left its natal home range and ended once individuals became
sedentary again. During dispersal, GPS collars recorded a fix every 4 hours and
they regularly transmitted data over the Iridium satellite system. After
collection, we converted GPS coordinates of dispersers (n = 4'169) to steps,
where each step represented the straight-line connecting two consecutive GPS
relocations \citep{Turchin.1998}. To ensure a regular sampling interval, we
removed fixes that were not successfully collected on the 4-hourly schedule (\(
\pm \) 15 minutes) and identified bursts of regularly sampled GPS fixes.

\subsection{Covariates}
We represented the physical landscape in our study area using a set of habitat
covariates, including water-cover, distance-to-water, woodland-cover, and
shrub/grassland-cover. Because water-cover greatly changes within and between
years in the Okavango Delta, we applied a remote sensing algorithm and obtained
weekly updated water-cover layers and corresponding distance-to-water layers
from MODIS satellite imagery (see \citealp{Wolski.2017} and Appendix A3 in
\citealp{Hofmann.2021}). Resulting layers thus temporally aligned with each
dispersal event, ensuring minimal temporal lag between observed movements and
corresponding water layers. We furthermore included a proxy for human influence,
rendering anthropogenic pressures stemming from human-density, agricultural
sites, and roads. All spatial layers were coarsened or interpolated to a target
resolution of 250 m by 250 m. A detailed description of the preparation of each
habitat covariate is provided in \cite{Hofmann.2021}.

>>>>

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}
\usepackage[colorinlistoftodos]{todonotes}

@@

%------------------------------------------------------------------------------
%  LONG VERSION
%------------------------------------------------------------------------------
% Importance of dispersal & connectivity and mow its measured
Dispersal of individuals is crucial for long-term species persistence and
depends on a sufficient degree of landscape connectivity. Thusfar, researchers
have investigated connectivity primarily using least-cost analysis and circuit
theory, although both methods make assumptions that are rarely met in reality.
Least-cost analysis assumes that animals move towards a known endpoint and are
knowledgeable about the least costly route to reach it. Circuit theory relies on
the assumption of a complete random walk without any directional persistence
along the dispersing route. Albeit these shortcomings can be overcome by
spatio-temporally explicitly simulating dispersing trajectories across the
landscape, a unified approach for such a simulation is lacking.

% What we do
Here, we present a simple three-step workflow to simulate dispersal and assess
connectivity starting from empirical GPS movement data. To exemplify its
application, we employ the workflow to dispersal data collected on the
endangered African wild dogs (\textit{Lycaon pictus}) in the Kavango-Zambezi
Transfrontier Conservation Area, the world’s largest transboundary conservation
area. In step one, we use integrated step selection functions to fit a
mechanistic movement model describing habitat and movement preferences of
dispersers. In step two, we apply the parameterized model to simulate individual
dispersal trajectories. In step three, we compress the simulated trajectories
into three complementary connectivity maps: a heatmap that highlights frequently
traversed areas, a betweenness map that pinpoints dispersal corridors, and a map
of interpatch connectivity, indicating the presence or absence of functional
links between habitat patches.

% Some results
Our results show that dispersing African wild dogs move with directional
persistence in the vicinity to waterbodies and preferrably through areas with
little human influence. This is reflected in the simulation with resulted in
several dispersal hotspots and corridors along waterbodies through areas with
little or no anthropogenic pressure. Connectivity between national parks inside
the KAZA-TFCA appears to be good, albeit the parks in Zambia and Zimbabwe are
comparably less well connected to the rest of KAZA-TFCA, suggesting the
necessity for additional protected areas that serve as ``stepping stones'' into
these regions.

@@

% Importance of Dispersal & Connectivity
\subsection{Importance of Connectivity \& Connectivity Models}
Dispersal of individuals is a vital process that allows species to maintain
genetic diversity \citep{Perrin.1999, Perrin.2000, Frankham.2002, Leigh.2012,
Baguette.2013}, rescue non-viable populations \citep{Brown.1977}, and colonize
unoccupied habitats \citep{Hanski.1999b, MacArthur.2001}. However, the ability
to disperse depends on a sufficient degree of landscape connectivity
\citep{Fahrig.2003, Clobert.2012}, making the identification and protection of
dispersal corridors that promote connectivity a task of fundamental importance
\citep{Nathan.2008, Doerr.2011, Rudnick.2012}. The identification of dispersal
corridors not only necessitates a comprehensive understanding of the factors
that limit dispersal, but also an appropriate model to estimate connectivity
\citep{Baguette.2013, Vasudev.2015, Hofmann.2021}. To date, the two most
commonly used connectivity models are least-cost path analysis (LCPA;
\citealp{Adriaensen.2003}) and circuit theory (CT; \citealp{McRae.2006,
McRae.2008}), both graph-based methods that quantify conductance of the
landscape based on habitat permeability \citep{Zeller.2012}. However, both
approaches rest on assumptions that appear unsuitable for dispersers, which is
why  simulating dispersal from individual-based movement models provides a
promising alternative approach to modeling connectivity \citep{Diniz.2019}.

% Issues with Both Methods
\subsection{Issues with Traditional Connectivity Models}
Traditional connectivity models make assumptions that are rarely met for
dispersers. LCPA, for instance, assumes that individuals move towards a
preconceived endpoint and choose a cost-minimizing route accordingly
\citep{Sawyer.2011, Abrahms.2017}. While this assumption may be fulfilled by
migrating animals, it is unlikely to hold for dispersers, as dispersers
typically move across unfamiliar territory and are therefore less aware of
associated movement costs and potential endpoints \citep{Koen.2014, Cozzi.2020}.
CT, on the contrary, posits that animals move according to a random walk,
entailing that autocorrelation between subsequent movements cannot be rendered
\citep{Diniz.2019}. For dispersers, however, autocorrelated movements are
regularly observed \citep{Cozzi.2020, Hofmann.2021}, implying that subsequent
relocations tend to fall onto a straight line. Moreover, graph-based methods are
unable to reflect the temporal dimension of dispersal, meaning that statements
about the approximate duration required to move between habitats impossible
\citep{Martensen.2017, Diniz.2019}.

% What about IBMMs?
\subsection{What about IBMMs?}
The shortcomings inherent to LCPA and CT can be overcome by simulating dispersal
trajectories from individual-based movement models (IBMMs) and by converting
simulated trajectories into meaningful measures of connectivity
\citep{Diniz.2019}). IBMMs allow to explicitly simulate how individuals move
across and interact with the encountered landscape \citep{Kanagaraj.2013,
Clark.2015, Allen.2016, Hauenstein.2019, Zeller.2020} and to render potential
interactions between movement behavior and habitat conditions. This shifts the
focus from a structural to a more functional view on connectivity
\citep{Tischendorf.2000, Kanagaraj.2013, Hauenstein.2019}. Furthermore,
simulations from IBMMs generate movement sequentially, making the temporal
dimension of dispersal explicit and allowing to model autocorrelation between
subsequent movements \citep{Diniz.2019}. Finally, simulations from IBMMs do not
enforce connections towards preconceived endpoints, thereby mitigating biases
arising from misplaced endpoints. Despite these advantages, a unified framework
to simulate dispersal and assess connectivity using IBMMs is lacking.

% Proposed Solution: Three-Step Workflow
\subsection{Proposed Solution: Three-Step Workflow}
Here, we draw on recent advancements in the fields of movement ecology and
network theory to propose a simple three-step workflow for simulating dispersal
movements and assessing landscape connectivity (\Cref{GraphicalAbstract}). In
step one, we combine GPS movement data of dispersing individuals with relevant
habitat covariates to fit a mechanistic movement model using integrated step
selection functions (ISSFs, \citealp{Avgar.2016}). ISSFs permit
simultaneous inference on the focal species' habitat kernel (i.e. habitat
preferences), its movement kernel (i.e. movement preferences/capabilities), and
potential interactions between the two \citep{Avgar.2016, Fieberg.2021}. In step
two,  we simulate individual dispersal trajectories using the movement model
parametrized in step one. Finally, in step three, we convert simulated
trajectories into three complementary connectivity maps, each highlighting a
different aspect of connectivity. The proposed set of maps includes (i) a
heatmap revealing areas that are frequently traversed by simulated dispersers
(e.g. \citealp{Hauenstein.2019, Zeller.2020}), (ii) a betweenness-map
delineating dispersal corridors and bottlenecks (e.g.
\citealp{BastilleRousseau.2018}), (iii) and a map of inter-patch connectivity,
depicting the presence or absence and intensity of use of specific connections,
and the average dispersal duration between corresponding habitat patches (e.g.
\citealp{Kanagaraj.2013}).

% Introduction of the study species
\subsection{Introduction of the Study Species}
The endangered African wild dog (\textit{Lycaon pictus}) represents a highly
mobile species whose persistence hinges on a  sufficient degree of landscape
connectivity. Once common throughout sub-Saharan Africa, this species has
disappeared from much of its historic range, largely due to human persecution,
habitat fragmentation, and disease outbreaks \citep{Woodroffe.2012}. Wild dogs
typically emigrate in single-sex coalitions \citep{McNutt.1996, Behr.2020}
capable of dispersing several hundred kilometers \citep{DaviesMostert.2012,
Masenga.2016, Cozzi.2020}. Although previous studies have investigated
connectivity using LCPA \citep{Hofmann.2021} or CT \citep{Brennan.2020}, a more
comprehensive and mechanistic understanding of connectivity is missing for this
species (but see \citealp{Creel.2020}). With fewer than 6,000 free-ranging wild
dogs remaining in fragmented populations \citep{Woodroffe.2012}, reliable
information on landscape connectivity is essential for the conservation of this
endangered carnivore.

% % Some Background on SSFs
% \subsection{Some Background on SSFs}
% To fit a movement model, we propose to employ the framework of step selection
% functions (SSFs; \citealp{Fortin.2005, Thurfjell.2014, Avgar.2016}). Originally,
% SSFs were developed to learn about habitat preferences of the focal species
% using GPS data \citep{Fortin.2005}. The method works by comparing covariates
% along realized steps (the straight-line connecting two consecutive GPS
% relocations; \citealp{Turchin.1998}) to the same covariates along randomly
% generated alternative steps \citep{Fortin.2005, Thurfjell.2014} using
% conditional logistic regression \citep{Fortin.2005, Muff.2020}. Recently,
% \cite{Avgar.2016} have introduced \textit{integrated} SSFs (ISSFs), which is a
% generalization of regular SSFs and requires that movement metrics are included
% in the corresponding regression model. This not only reduces biases in estimated
% preferences, but also permits simultaneous inference on a species' habitat
% kernel (i.e. habitat preferences), its movement kernel (i.e. movement
% preferences/capabilities), and potential interactions between the two kernels
% \citep{Avgar.2016, Fieberg.2021}. Importantly, a model that was fitted using
% ISSFs resembles a fully mechanistic IBMM, based on which movement can be
% simulated \citep{Avgar.2016, Signer.2017}. Previously, ISSFs have been applied
% to simulate utilization distributions of resident animals \citep{Avgar.2016,
% Signer.2017}, yet the method remains underutilized for the simulation of
% dispersal and detection of movement corridors.

% % Some Background on Connectivity "Metrics"
% \subsection{Some Background on  Connectivity ``Metrics'' }
% Irrespective of the chosen modeling framework, simulated dispersal trajectories
% need to be processed into measures of connectivity. To date, the most common
% practice to translate simulations into connectivity measures is to compute
% inter-patch connectivity between distinct habitats \citep{Diniz.2019}.
% Inter-patch connectivity is a simple ratio between the number of dispersers
% successfully moving between habitat patches and the total number of simulated
% individuals from each patch (e.g. \citealp{Kanagaraj.2013}). As an alternative,
% some researchers have tallied overlapping trajectories into what might be called
% ``heatmaps'', i.e. maps that depict the frequency at which each spatial unit in
% the study area was traversed by simulated dispersers (e.g. \citealp{Peer.2008,
% Hauenstein.2019, Zeller.2020}). More recently, \cite{BastilleRousseau.2018}
% proposed a method to convert trajectories into a network, so that metrics from
% network theory can be computed. With regards to connectivity, the betweenness
% metric seems promising, as it indicates the number of shortest paths going
% through each node in the network and therefore highlights areas of exceptional
% importance for connecting different regions in the study area. Because each of
% the three metrics metrics focuses on a very different aspect of landscape
% connectivity, we propose to view them as complementary and to use them in
% concert when assessing connectivity.

@@

% Case Study
\subsection{Case Study}
In this study, we illustrate the proposed three-step workflow
(\Cref{GraphicalAbstract}) to assess landscape connectivity for African wild
dogs within the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA), the
world's largest transboundary conservation area. To this end, we use GPS
movement data of 16 dispersing wild dogs originating from a free-ranging
population in northern Botswana to parametrize a mechanstic movement model,
which we then employed to simulate 80,000 dispersal trajectories across the
landscape of the KAZA-TFCA. We anticipated that simulations based on our
three-step workflow would overcome several of the above highlighted conceptual
shortcomings of traditional connectivity models and provide a more comprehensive
view of landscape connectivity.

\section{Methods}
\subsection{Study Area}
The study area centered at -17\degree 13'9''S, 23\degree 56'4''E
(\Cref{StudyArea}a) is located in southern Africa (\Cref{StudyArea}a) and spans
more than 1.3 Mio. km\textsuperscript{2}, encompassing the entire KAZA-TFCA
(\Cref{StudyArea}b). The KAZA-TFCA comprises parts of Angola, Botswana, Namibia,
Zimbabwe, and Zambia and hosts diverse landscapes; ranging from savannah to
grassland and from dry to moist woodland habitats. In its center lies the
Okavango Delta, a dominant hydrogeographical feature and the world's largest
flood-pulsing inland delta. Although large portions of the KAZA-TFCA are part of
designated national parks and other protected areas, considerable human
influence from roads, agricultural sites, and settlements remains.

\begin{figure}[h]
  \begin{center}
    \begin{tikzpicture}
        \node[anchor=south west,inner sep=0] (image) at (0,0,0) {
        \includegraphics[width=\textwidth]{99_StudyArea.png}
        };
        \begin{scope}[x={(image.south east)},y={(image.north west)}]
            % % next four lines will help you to locate the point needed by forming a grid.
            % % comment these four lines in the final picture.
            % \draw[help lines,xstep=.1,ystep=.1] (0,0) grid (1,1);
            % \draw[help lines,xstep=.05,ystep=.05] (0,0) grid (1,1);
            % \foreach \x in {0,1,...,9} { \node [anchor=north] at (\x/10,0) {0.\x}; }
            % \foreach \y in {0,1,...,9} { \node [anchor=east] at (0,\y/10) {0.\y};}
            % % upto here
            \draw[densely dotted, blue] (0.169, 0.222) -- (0.364, 0.955);
            \draw[densely dotted, blue] (0.169, 0.157) -- (0.364, 0.074);
        \end{scope}
    \end{tikzpicture}
    \caption{Illustration of the study area located in southern Africa. (a) The
    study area was confined by a bounding box spanning the entire KAZA-TFCA and
    comprised parts of Angola, Namibia, Botswana, Zimbabwe, and Zambia. (b)
    The KAZA-TFCA represents the world's largest terrestrial conservation area
    and covers a total of 520'000 km\textsuperscript{2}. Its main purpose is to
    re-establish connectivity between already-existing national parks (dark
    green) and other protected areas (light green).}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{Data Collection and Preparation}
\subsubsection{GPS Data}
We collected GPS movement data on 16 dispersing wild dogs (7 females and 9
males) between 2011 and 2019 from a free-ranging population in northern Botswana
(details on the data collection can be found in \cite{Cozzi.2020} and
\cite{Hofmann.2021}). Because behavior during dispersal is more pertinent to
landscape connectivity than behavior during residence \citep{Elliot.2014,
Abrahms.2017}, we discarded data collected when individauls were resident. We
determined the exact timing of emigration and settlement using direct field
observations and through visual inspection of the net squared displacement (NSD)
metric. NSD measures the squared Euclidean distance of a GPS relocation to a
reference point \citep{Borger.2012}, which we set to the center of each
dispersers' natal home range. Thus, dispersal was deemed to have started once an
individual left its natal home range and ended once the NSD metric ramined
constant, indicating settlement. During dispersal, GPS collars recorded a fix
every 4 hours and they regularly transmitted data over the Iridium satellite
system. To ensure regular timespans between fixes, we removed any fix that was
not obtained on the aspired 4-hourly schedule (\( \pm \) 15 minutes). We then
converted the remaining fixes (n = 4'169) into steps, where each step
represented the straight-line movement between two consecutive GPS relocations
\citep{Turchin.1998}.

\subsubsection{Habitat Covariates}
We represented the physical landscape in our study area by the habitat
covariates \textit{water-cover, distance-to-water, woodland-cover,
shrub/grassland-cover, and human-influence}. To reflect the seasonality of water
in the study area, we obtained weekly updated layers for water-cover and
corresponding layers for distance-to-water from MODIS satellite imagery using a
remote sensing algorithm \citealp{Wolski.2017, Hofmann.2021} that is implemented
in the \textsf{floodmapr} package which is available on GitHub
(\url{https://github.com/DavidDHofmann/floodmapr}). For each step, we thus used
the temporally associated water layers, reflecting prevailing conditions at the
time when the step was realized. To ensure a consistent resolution across
habitat covariates, we coarsened or interpolated all layers to a resolution of
250 m x 250 m. A more detailed description of how we prepared each habitat
covariate is provided in \cite{Hofmann.2021}.

Besides habitat covariates, we computed movement metrics that we used as
movement covariates in our ISSF analysis \citep{Avgar.2016, Fieberg.2021}.
Movement metrics were calculated for each step and included the step length
(\textsf{sl}), its natural logarithm (\textsf{log(sl)}), and the cosine of the
relative turning angle (\textsf{cos(ta)}).  Moreover, we created the binary
variable \textsf{LowActivity} indicating whether a step was realized during
periods of low wild dog activity (09:00 to 17:00 local time) or high wild dog
activity (17:00 to 09:00 local time, \citealp{Cozzi.2012}). We performed all
data preparations, spatial computations, and statistical analysis in R, version
3.6.6 \citep{R.2020}. Some helper functions were written in {\tt C++} and
imported into {\tt R} using the {\tt Rcpp} package \citep{Eddelbuettel.2011,
Eddelbuettel.2013}. We provide annotated R-codes to reproduce the three-step
workflow using simulated data through GitHub
(\url{https://github.com/DavidDHofmann/DispersalSimulation}). In addition, codes
for the African wild dog case study are available through an online repository.

\subsection{Step 1 - Movement Model}
We used ISSFs to parametrize a mechanistic movement model for dispersing wild
dogs \citep{Avgar.2016}. More specifically, we paired each realized step with 24
random steps, so that a realized step plus its 24 random steps formed a stratum
that received a unique identifier. As suggested by \cite{Avgar.2016}, we
generated random steps by sampling random turning angles from a uniform
distribution (\(-\pi, +\pi\))) (which is equivalent to a von Mises distribution
with concentration parameter \(\kappa = 0\)) and step lengths from a gamma
distribution that was fitted to realized steps (scale \(\theta\) = 6'308 and
shape \(k\) = 0.37)). Along each step, we extracted and averaged values from the
habitat covariate layers using the {\tt velox} package \citep{Hunziker.2021} and
calculated the movement metrics \textsf{sl}, \textsf{log(sl)}, and
\textsf{cos(ta)} for each realized and random step. To facilitate model
convergence, we standardized all continuous covariates to a mean of zero and a
standard deviation of one. Since correlation among covariates was low (\(|r| <
0.6\); \citealp{Latham.2011}), we retained all of them for modeling.
